---
title: "An exploratory analysis of the PRIO"
date: 2016-11-20
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Intro

This is an overview of the dataset released by [UCDP - Uppsala Conflict Data Program](http://www.ucdp.uu.se/) containing individual events of organized violence (phenomena of lethal violence occurring at a given time and place).

## Thanks
Sundberg, Ralph, and Erik Melander, 2013, “Introducing the UCDP Georeferenced Event Dataset”, Journal of Peace Research, vol.50, no.4, 523-532
Croicu, Mihai and Ralph Sundberg, 2015, “UCDP GED Codebook version 2.0”, Department of Peace and Conflict Research, Uppsala University

# Get and prepare data

Database is downloaded in R Data format from the url [http://ucdp.uu.se/downloads/ucdpprio/ucdp-prio-acd-4-2016.RData.zip](http://ucdp.uu.se/downloads/ucdpprio/ucdp-prio-acd-4-2016.RData.zip)

```{r}
load("~/Downloads/ged50.Rdata")
library(tidyverse)
db <- tbl_df(ged50)
```

At the same location there is a codebook in pdf format that covers extensively all variables contained in the dataset.

An overview of the variables:

```{r}
db %>% glimpse()
```

# Some analysis

## Events by year
```{r}
db %>% 
 ggplot(aes(x = year)) +
  stat_count() +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Year",
       y = "Count",
       title = "Events by year")
```

## Killed by year
```{r}
db %>% 
  group_by(year) %>% 
  summarise(nkill = sum(best_est)) %>% 
  ggplot(aes(x = year, y = nkill)) +
  geom_point() +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Year",
       y = "Number of killed",
       title = "Number of deaths by year")
```
